---
title: "Six decades of deforestation and forest fragmentation in Madagascar"
bibliography: bib/biblio.bib
csl: bib/conservation-letters.csl
output: 
  html_document:
    fig_caption: yes
  pdf_document:
    fig_caption: true
    keep_tex: true
---

```{r load_data}
load("deforestmap.rda")
```

#### Ghislain Vieilledent<sup>*,1,2</sup>, Clovis Grinand<sup>3</sup>, Fety A. Rakotomalala<sup>3</sup>, Rija Ranaivosoa<sup>5</sup>, Jean-Roger Rakotoarijaona<sup>5</sup>, Thomas F. Allnutt<sup>4</sup>, and Frédéric Achard<sup>1</sup>

(*) /Email: ghislain.vieilledent@cirad.fr, /Phone: +39.329.457.2273

1. Joint Research Center of the European Commission, Bio-economy Unit (JRC.D.1), I-21027 Ispra (VA), ITALY
2. Cirad, UPR Forêts et Sociétés, F-34398 Montpellier, FRANCE
3. ETC Terra, F-75016 Paris, FRANCE
4. Wildlife Conservation Society, 101 Antananarivo, MADAGASCAR
5. Office National de l'Environnement, 101 Antananarivo, MADAGASCAR

## Summary

1. xxx

2. xxx

**Keywords**: biodiversity, climate-change, deforestation, Landsat, protected areas, Madagascar, tropical forest, tree cover.

**Potential journals** (2015 IF): Conservation Letters (7.126), Conservation Biology (4.257), Environmental Research Letters (4.134), Biological Conservation (3.985), Landscape Ecology (3.657), Ambio (2.555), Environmental Conservation (2.235), Biotropica (1.944), Tropical Conservation Science (1.55).

## 1. Introduction

Tropical forests provide important ecosystem services at the global scale, such as biodiversity conservation and climate regulation [@Costanza1997] but also many at the regional scale for local populations. At the local scale, tropical forests supply clean water, can preserve populations from flood [@Bradshaw2007] and provide wood for construction and fuel and other non-woody products for food and health. On the island of Madagascar, three types of tropical forests, covering about 9 millions of hectares (15% of the country area) can be found: the moist forest in the East, the dry forest in the West and the spiny forest in the South [@Vieilledent2016]. Among all the ecosystem services cited previously, Malagasy forests are particularly important for the unique biodiversity they shelter, both in term of number of species and endemism in many taxonomic groups [@Goodman2005]. Deforestation threatens species survival and diminishes biodiversity by destroying forest habitat, creating forest fragments too small to maintain viable populations and increasing 'edge effects' at forest/non-forest interfaces (Harris 1984). Edge effects due to fragmentation typically affect an area several times larger than the forest destruction itself (Harris 1984; Skole & Tucker 1993), affecting micro-meteorology over short distances (Kapos 1989), and increasing exposure to damaging winds (Ferreira & Laurance 1997), fire
frequency (Cochrane 2001), and access for livestock, other non-forest animals and hunters (Enserink 1999; Cullen et al. 2000). To assess the impact of tropical deforestation on biological diversity, not only the area deforested, but also the isolation of forest patches and the area of edge habitat must be determined.

Accurate maps of forest cover and forest cover change are necessary to monitore carbon emissions and loss of biodiversity associated to deforestation, assess the efficiency of present conservation strategies [@Eklund2016], and implement new strategies for the future [@Vieilledent2013a, @Vieilledent2016]]. Unfortunately, accurate and cloud-free forest cover maps are not available for Madagascar for the last decade (2005-2015). In their paper untitled _"Fifty years of deforestation and forest fragmentation in Madagascar"_, @Harper2007 published maps of forest cover change for Madagascar from 1950s to 2000 (for the years _c._ 1953, _c._ 1973, 1990 and 2000). The 1950s forest map was derived from the visual interpretation of aerial photographies and gives a coarse representation (scale 1/1,000,000) of the forest extent in Madagascar for the 1950s. Forest maps for the years _c._ 1973, 1990 and 2000, were obtained from a supervised classification of Landsat satellite images at 30m and can be used to derive much more accurate estimates of the forest cover at these three dates. Notheless, maps provided by @Harper2007 were not free of clouds (11 244 km2 of unknown cover type for the year 2000 for example). Using the same supervised classification approach as in @Harper2007, more recent maps have been released for the periods 2000-2005-2010 by national institutions, with the technical support of international environmental NGOs [@MEFT2009, @ONE2013]. Another recent map using more advanced statistical tool for classification (the Random Forest classifier, see @Grinand2013 and @Rakotomalala2015) has been released for the periods 2005-2010-2013 [@ONE2015]. However, these maps might (i) be too old to give recent estimates of deforestation [@MEFT2009, @ONE2013], (ii) figure large areas of clouds (images with more than 20% of clouds in 2010 for the map provided by [@ONE2013]), or (iii) show large errors of classification in specific areas, especially in the dry and spiny forests for which the spectral answer varies strongly due to deciduousness associated with the higher seasonality than in moist forests (Kappa<0.6 and overall accuracy<0.8 for the dry and spiny forests in @ONE2015). Moreover, deriving such forest maps from supervised classification requires a huge amount of time and work, especially regarding image selection to reduce cloud cover and visual interpretation of the numerous observation polygons used to train the classification algorithm. In addition, visual interpretation might be subjective and is potentially a source of errors. All this work of image selection and visual interpretation will have to be repeated to produce new forest maps in the future.  

Global forest products have also been published recently and could be used to derive recent forest cover map at the national scale for Madagascar. @Kim2014 produced a global Landsat-based forest-cover change map from 1990 to 2000 but their approach was not reproduced to obtain more recent maps of forest-cover, after 2000. Also, the approach used in @Kim2014 did not allow to correctly map the forest in the dry and spiny ecosystems. Another global study is the work by @Hansen2013 who identified annual tree cover, forest loss and forest gain from 2000 to 2012. Meanwhile, results have been updated and are available up to the year 2014 at http://earthenginepartners.appspot.com/science-2013-global-forest. One problem with using the @Hansen2013 product to map forest cover is that a tree cover threshold must be defined to transform the tree cover data into a forest/non-forest map. Identifying such threshold is not straightforward as forests cannot be defined only by a tree cover value. Also, the tree cover threshold might strongly vary between forest types, mean tree cover values being substantially lower for dry forests than for moist forests. Also, a tree cover gives no information on land-use. Therefore, tree plantations such as oil palm or eucalyptus plantations cannot be differenciated from natural forests, which provide very different ecosystem services particularly in terms of biodiversity [@Tropek2014]. Thus, using @Hansen2013 alone can thus only give a first-approach estimate of the forest cover. But, tree cover is not the only information provided by @Hansen2013. The forest loss, i.e. the complete loss of forest tree cover (which shifts from a positive value over 10% to zero) is also reported.

In this study, we present a simple approach combining maps from @Harper2007 and products from @Hansen2013 to derive annual and cloud-free forest-cover change map for Madagascar for the period 2000-2014. Basically, we used the forest map provided by @Harper2007 for the year 2000 (defining the land-use) with the tree cover loss product provided by @Hansen2013 that we apllied only in forest areas. Following the work done by Harper2007, we studied in details the change in deforestation rates and the evolution of forest fragmentation on six decades from _c._ 1953 to 2014. We showed that despite all the conservation efforts done in Madagascar since the mid 1980s, deforestation has increased in speed since 2010 and that fragmentation has dramatically increased leading to few remaining intact forest landscapes. The approach described in this study could be easilly extended to all tropical countries with a forest map obtained at one date between 2000 and 2014, and could be repeated in the future to obtain future forest-cover change maps, as and when the @Hansen2013 product is updated.

## 2. Materials and Methods

### 2.1 Historical deforestation maps

We derived 30m cloud-free forest maps for Madagascar for the years 1990, 2000, 2010 and 2014 combining the 1990 and 2000 forest maps by @Harper2007 and the 2000-2014 tree percent cover map by @Hansen2013. On Harper's map, ~200,000 ha clouds are present over the ~4.2 million ha moist forest ecoregion (4.8% of clouds). To remove these clouds, we used the 2000 cloud-free tree percent cover map provided by @Hansen2013 (also at 30~m resolution) and we chose a threshold of 75% of tree cover to decide whether to replace cloud pixels with forest or non-forest pixels [@Achard2014]. We thus obtained a cloud-free year 2000 forest map for Madagascar. From this map, using tree cover loss data from 2000 to 2014 by @Hansen2013, we obtained cloud-free forest maps in 2010 and 2014 at 30~m resolution. Finally, we corrected the cloudy 1990 Harper forest map assuming that if forest was present in 2000, it was also present in 1990. Doing so, all clouds were removed and we obtained a cloud-free 1990 forest map for Madagascar. These maps are freely available at http:/bioscenemada.cirad.fr/forestmaps. On the basis of the forest maps, we computed the mean annual deforestation intensity (in hectares of forest and in percentage) for the five period of time from 1990 to 2014.

Regarding the gains: we did not consider regrowth:
- few potential regrowth in Madagascar.
- regrowth are usually more easilly re-burnt for agriculture on the long term.
- young secondary forests do not provide the same ecosystem services (in terms of biodiversity and carbon storage) as old growth natural forests.
- Harper intro: "we saw little evidence of forest regeneration in Madagascar".

### 2.2 Validation

We estimated the accuracy of our approach combining Harper's and Hansen's products to derive cloud-free forest-cover change maps after the year 2000 in Madagascar. To do so, we used a forest-cover change map obtained from a regional study on the humid ecoregion on the period 2005-2013 [@Rakotomalala2015]. @Rakotomalala2015 performed a supervised classification of Landsat images at multiple dates using the Random Forest classifier [@Grinand2013]. The map by @Rakotomalala2015 is considered rather accurate on the moist ecoregion with an external validation on nearly 10,854 visual interpretation points leading to rather high Kappa and overall-accuracy values (0.79 and 0.89 respectively). Following our methodology, we produced a deforestation map on the same period of time (2005-2013) and compared the forest-cover (in ha) and mean annual deforestation rates (in ha and % per year) between the two maps.

### 2.3 Fragmentation

We report an analysis of the evolution of forest fragmentation based on the forest-cover maps for Madagascar for the years _c._ 1953, _c._ 1973, 1990, 2000, 2010 and 2014. We used the method developed by @Riitters2000 which uses a moving window to characterize the fragmentation around each forested pixel. Computation were done using function `r.forestfrag` from GRASS GIS software [@Neteler2008]. Six categories of fragmentation are identified from the amount of forest and its occurrence as adjacent forest pixels: interior, perforated, edge, transitional, patch, and undetermined. We used a moving window of 7x7 pixels (4.41 km<sup>2</sup>). Using this window size, forest edge had a width of about 90 m [@Riitters2000]. The interior category can be interpreted as the most intact forest (sensus @Potapov2017). We reported the area of forest in each fragmentation category for the six years and analysed the evolution of the fragmentation with time.

## 3. Results

### 3.1 Validation

Difference might be due to post-processing of @Rakotomalala2015 map.

### 3.2 Evolution of forest cover and deforestation intensity

### 3.3 Evolution of forest fragmentation with time



## 4. Discussion

### 4.1 Evolution of deforestation and forest fragmentation in Madagascar on the last 60 years

### 4.2 Interest of such an approach and limitations

- simple
- easilly reproducible
- no need to define a tree cover threshold to 

## 5. Tables

```{r comp, results="asis"}
tab.1 <- comp_moist.df
mean.diff <- round(100*(sum(tab.1$BSM)-sum(tab.1$PERR_FH))/sum(tab.1$PERR_FH))
names(tab.1) <- c("Year","Vieil (ha)", "Rakot (ha)", "Rakot_NA (ha)", "ad_Vieil (ha/yr)", "ad_Rakot (ha/yr)", "theta_Vieil (%/yr)", "theta_Rakoto (%/yr)", "diff.area (%)")
tab.1 <- tab.1[,c(1:4,9,5:8)]
kable(tab.1)
```

Table 1: **Comparison of the forest cover and deforestation rates between our study and the study by @Rakotomalala2015**. This comparison is done only for the humid ecoregion for the periods 2005-2010-2013, which is the extent and time-periods of the @Rakotomalala2015 map. @Rakotomalala2015 map has been validated using 10,854 independant visual interpretation points leading to a rather high accuracy (Kappa=0.79 and overall-accuracy=0.89). Our study (suffix _"Vieil"_) estimated a forest cover greater by about `r mean.diff`% compared to @Rakotomalala2015 (suffix _"Rakot"_). But the difference (_"diff.area"_) did not vary much between dates (between 3.56 and 6.18%). Our study also estimated higher annual deforestation (_"ad"_) and deforestation rates (_"theta"_) but the same trend for the deforestation was observed between the two studies, with a doubling of the deforestation for the period 2010-2013 compared to 2005-2010. The differences between the two studies could be attributed to the forest definition which might be different in the two studies and to the post-processing in @Rakotomalala2015 map.

```{r defor, results="asis"}
tab.2 <- defor.df
names(tab.2) <- c("Year","Forest (ha)", "Annual defor. (ha/yr)", "theta (%/yr)")
kable(tab.2)
```

Table 2: **Evolution of the forest cover and deforestation intensity from 1950s to 2014 in Madagascar**. Madagascar has lost 37% of its natural forest in 40 years between _c._ 1973 and 2014 (and roughly 56% in 60 years since _c._ 1953). If the deforestation has progressively slowed down from _c._ 1953 to 2010 (from 1.76 to 0.59%/yr), most likely due to a decrease of available arable lands under the remaining forests and to a decrease of forest accessibility, the deforestation has recently increased, with a doubling deforestation on the period 2010-2014 compared to 2000-2010. This increase could be attributed to both political instability (political crisis in 2009) and to an explosion of the demography (according to the @UN2015, population has increased from about 16M inhabitants in 2000 to 24M in 2014). Natural forest cover 8.9 Mha in 2014 and current (period 2000-2014) mean annual deforestation rate is of about 70,000 ha/yr.   

```{r frag, results="asis"}
tab.3 <- frag.df
names(tab.3) <- c("Year","Forest (ha)", "patch (%)", "transitional (%)", "edge (%)", "perforated (%)", "interior (%)", "NA (%)")
kable(tab.3)
```

Table 3: **Evolution of the forest fragmentation from 1950s to 2014 in Madagascar**. Forest fragmentation has progressively increased from _c._ 1953. The percentage of forest belonging to the interior category (the most intact forest) has fallen from 94 in _c._ 1973 to 48% in 2014. More than 15% of the forest belong to the patch and transitional categories (isolated small forest patches) in 2014. This increase in forest fragmentation has a great impact on both carbon emissions [@Putz2014] and biodiversity loss [@Gibson2013, @Murcia1995].

## 6. Figure captions

## 7. Figures

## Acknowledgements

This study is part of the BioSceneMada project (http://bioscenemada.cirad.fr) and the ReCaREDD project (http://roadlessforests.org). The BioSceneMada project is funded by FRB (Fondation pour la Recherche sur la Biodiversité) and the FFEM (Fond Français pour l'Environnement Mondial) under the project agreement AAP-SCEN-2013 I. The ReCaREDD project is funded by the European Commission.

## Data availability statement

All the data and codes used for this study are made publicly available in the `deforestmap` GitHub repository (https://github.com/ghislainv/deforestmap.git). The results and the manuscript are fully reproducible running the R script `deforestmap.R` located inside the `deforestmap` repository.

## References
